Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-04 paper/paper.qmd
commit changes to paper.qmd

## Measurement
Voters’ opinions are multifaceted and can be influenced by numerous factors, including media coverage, recent political events, personal beliefs, and social pressures. Each voter’s decision is shaped by a unique mix of these factors, which can change frequently, especially close to an election. Pollsters start by converting the complex, multi-layered opinions in voters' heads into a simple polling response. This is the first challenge: each voter’s thoughts might not fit neatly into options like “Trump,” “Harris,” or “Undecided.” People may lean toward one candidate but have reservations, or may change their views over time. By asking for a single answer, the poll essentially “forces” an opinion that may not fully capture the voter’s internal conflict.

Once collected, these individual responses are combined into percentages of support. This aggregation abstracts further, treating all “Trump” responses equally, regardless of how confident or reluctant the voter might have been. In effect, the complexity of each opinion is smoothed over into a uniform data point, with many variations of opinions lost in the process. Finally, these support percentages are fed into a model to make a predictive statement about the election outcome. The model’s goal is to determine who is likely to win, based on the assumption that the aggregated opinions in our data reflect real-world voter behavior on Election Day. However, this step relies heavily on simplification, assuming that the snapshot of opinions captured in the polls will hold steady and that all voters will follow through on their expressed preferences. The raw polling data gives us percentages of support for each candidate in each state. To measure our outcome variable (the likelihood of winning), we convert these percentages into a binary outcome in our model (predicting Trump or Harris as the winner based on who has higher support in each state).  Our model uses these support percentages to generate a probability of each candidate winning. This transformation helps translate fluctuating polling data into a clear prediction of the likely winner, but it also involves a level of simplification, as it assumes these support percentages accurately reflect voting intentions.

Using state as a predictor variable, which influences our outcome variable, assumes that all responses within a state are comparable and that regional differences are consistent across the state. However, each state has internal diversity—urban vs. rural areas, differences in income levels, or cultural values—that the variable “state” may not fully capture. This simplification is necessary for the model, but it means that nuanced, within-state variations in opinion are lost in the abstraction. Another predictor variable, end_date is also measured, it is designed to capture how voter sentiment changes over time, especially as the election approaches. By using the end date of each poll, the model can recognize patterns or shifts in support due to campaign events, news coverage, or debates. While “end_date” provides temporal context, it simplifies complex, often continuous changes in opinion into a discrete snapshot. This abstraction assumes that voter sentiment is relatively stable between poll dates, which may not reflect reality. For example, an event occurring immediately after a poll’s end date could dramatically shift opinions that the model wouldn’t capture until the next poll. Additionally, the importance of time in influencing voter sentiment is assumed to be similar across states, but certain states might react differently to events depending on their local context or demographic makeup.

@@

- **'State'**: Indicates the state where the polling took place, critical for understanding regional support dynamics.
- **'Polling Date'**: The date when the poll was conducted, which helps in analyzing trends over time.
- **'Pollster'**: The organization conducting the poll, providing insight into the reliability and methodology of the polling data.
- **'N'umeric Grade'**: A score assigned to each poll based on its methodology and historical accuracy, which assists in filtering out less reliable polls.
- **'Transparency Score'**: A measure indicating how transparent the pollster is regarding their methodology, further enhancing data credibility.
- **'Candidate Name'**: Identifies the candidate being polled, specifically focusing on Donald Trump and Kamala Harris.
- **Percentage of Support (pct)**: Represents the percentage of respondents supporting each candidate, which is the primary outcome variable of interest.

@@

## Model Overview
The Bayesian Beta regression models constructed in this paper aim to predict the probability of victory for Donald Trump and Kamala Harris in the 2024 U.S. Presidential Election by analyzing polling data of the 7 battleground states. One Bayesian Beta regression model is constructed for each candidate, and each of them estimates the proportion of support for each candidate based on polling data, accounting for both polling trends over time, and state-level differences in voter sentiment. These models allow us to evaluate how support for each candidate fluctuates over time in key states, and predict potential electoral outcomes.

One of the goals of our model is to analyze changes in voter support for each candidate over time, as Election Day approaches. By using end_date_num (days the earliest poll date) as a predictor with natural splines, the model captures non-linear trends in support. The time-based analysis provides us with information about how voter preference may shift due to campaign events, news cycles, and other influential factors in the months leading up to the election.

Another goal of our model is to predict the potential winner of the Presidential Election. We recognize that battleground states have unique political landscapes, so we include state as a random effect in each model. This allows each of the 7 battleground states to have a unique baseline level of support for each candidate, capturing regional differences in voter preferences. By incorporating state-level effects, the model can show variations in support across different states, revealing the potential winning candidate for the presidential election. Further background details, model specifications, and diagnostics are included in [Appendix -@sec-model-details].

@@

## Model Setup
Our models were ran in R [@citeR] using the 'rstanarm' package of @rstanarm.

@@

## Uncertainty Beyond Data and Modeling
Beyond data and modeling, several external factors can influence the actual outcome of the election. Recency bias is one of the external factors, and it can significantly shape public perception and polling responses. Many people expect Trump to win due to his unexpected victory in 2016 and his narrow loss in 2020, despite lagging in the polls. This bias reflects a tendency to focus on recent, high-profile events while overlooking broader historical trends. For instance, in 2012, Barack Obama not only won but outperformed polling predictions, a fact that may be forgotten amidst recent election narratives [@discussion-article3]. Such a bias can affect both how people respond to polls and how pollsters interpret results, potentially leading to an overemphasis on recent events. Given that our model depends on polling data, any systematic bias from these recency expectations can skew predictions, making it challenging to account for potential polling errors.

Social desirability bias and non-response bias add further layers of uncertainty, particularly regarding Trump’s support base. Some voters may be hesitant to openly express support for a conservative candidate due to social stigma, leading to an under-reporting of Trump support in polls. This effect may be stronger in regions where conservative views are less socially accepted. Additionally, Trump’s supporters often include individuals with lower social trust and civic engagement, making them less likely to participate in polls altogether [@discussion-article3]. If specific groups systematically avoid participation, polling data may under-represent these voters, potentially underestimating Trump’s support in our model. Non-response bias is difficult to quantify, introducing additional uncertainty to our predictions.

Harris, on the other hand, may face challenges like the Bradley effect and gender-related bias. As a Black female candidate, Harris could encounter the Bradley effect, where some voters indicate indecision rather than openly stating that they will not support a minority candidate. Historically, undecided voters have leaned against Black candidates in the actual vote count, despite neutral polling responses. Additionally, the 2016 election demonstrated that undecided voters could heavily swing against female candidates, as seen with Hillary Clinton [@discussion-article3]. This trend could similarly impact Harris’s results if undecided voters ultimately decide against her. Consequently, our model may overestimate Harris’s support based on polling data alone, leading to potential inaccuracies in prediction.

These biases are inherently challenging to measure or adjust for, as they reflect deeper societal attitudes and individual decision-making patterns. While pollsters may attempt to correct for non-response or social desirability bias, such adjustments are often based on assumptions that cannot fully capture the complex social dynamics at play. Thus, our model’s predictions should be interpreted with caution, recognizing that these biases may cause unexpected shifts in voter behavior that go beyond what data-driven forecasts can capture.

@@

- **Pollster Ratings and Credibility**: Polls from different organizations often vary in reliability, as some pollsters have more accurate or less biased methodologies. By integrating pollster ratings, the model could assign weights to different polls based on the historical accuracy and reputation of each organization. This adjustment would allow the model to downplay the influence of potentially biased or less credible polls, particularly in competitive battleground states, where small differences in predicted support can significantly affect the outcome. This approach would lead to a more calibrated model that accounts for poll variability and thus provides a more reliable prediction.
- **Timing and Recency of Polls**: Voter sentiment is not static, and the timing of polls can significantly impact their relevance. As the election date nears, polls generally become more reflective of the final outcome. Incorporating a temporal weighting scheme that prioritizes recent polls could enhance the model’s responsiveness to last-minute shifts in voter preferences, which often occur due to campaign events, debates, or emergent political issues. This weighting approach could also help mitigate the risk of using outdated data that no longer accurately reflects current voter sentiment.
- **Economic and Political Context**: The political landscape is shaped by various external factors, such as economic indicators (e.g., unemployment rates, inflation) and prominent political events. By incorporating economic data or major political issues, such as health care or immigration policies, the model could capture contextual variables that influence voter sentiment. Such contextual factors could be especially useful in a Bayesian framework, allowing us to adjust prior distributions based on known historical influences on voter behavior.
- **Inclusive of National/Popular votes**: our model has exclusive focus on the seven battleground states—Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin. By narrowing the dataset to only these critical swing states, our model omits data from the rest of the nation, including states with predictable voting patterns (solidly Democratic or Republican) and states outside the battleground category. This omission is significant as it restricts the model’s capacity to capture the broader national voting landscape and limits the ability to provide a holistic prediction of the 2024 U.S. Presidential Election outcome. In particular, the exclusion of national popular vote data limits our model’s understanding of overall voter sentiment across the country. While battleground states are pivotal in determining the electoral outcome due to the Electoral College system, popular vote trends from non-battleground states can still reveal important national shifts in sentiment that may indirectly influence battleground dynamics. For example, a nationwide trend towards or away from a candidate could suggest emerging political momentum or reflect the impact of major events that also resonate within battleground states. Without incorporating this popular vote data, our model may overlook broader trends that could subtly shape battleground results, especially if voters in battleground states are influenced by national sentiment.

@@

# Additional Model details {#sec-model-details}
## Model Justification - Further Details
The Bayesian Beta Regression model with a logit link function is particularly suited for our prediction of the 2024 U.S. Presidential Election due to its alignment with the nature of our outcome data and the goals of our analysis. Our primary outcome variable is the percentage of support for each candidate in the seven battleground states, which we scaled to a proportion between 0 and 1. Since support percentages are naturally bounded between 0% and 100%, modeling these data with traditional linear regression presents challenges, as linear regression assumes an unbounded continuous outcome. The Beta distribution is ideal for this type of data because it is defined on the interval (0, 1), allowing us to model proportions without the risk of generating unrealistic values outside this range. Unlike a linear model, which could predict support percentages exceeding 100% or falling below 0%, the Beta distribution respects the boundaries of our data.

To effectively capture the dynamics of voter support, we utilized natural splines with the variable end_date_num, which represents the number of days since the earliest poll date. This choice allows the model to adapt to non-linear trends in voter support that can fluctuate due to various campaign events, debates, or shifts in the political landscape. By specifying four degrees of freedom for the natural splines, we enable the model to flexibly capture gradual changes in voter sentiment while mitigating the risk of overfitting to noise or abrupt fluctuations in the data. This flexibility is particularly critical in electoral polling, where opinions may shift slowly or rapidly in response to unpredictable external events. Alternative models, such as standard logistic regression without splines, would fail to capture these complex changes, potentially leading to a less accurate depiction of how support varies over time.

A unique aspect of our analysis is the focus on battleground states, which exhibit distinct political dynamics that can significantly influence election outcomes. Incorporating state as a random effect provides the flexibility necessary to estimate state-specific baseline support levels for each candidate while still allowing us to identify overall trends across all states. This approach is more effective than treating state as a fixed effect, as it avoids overparameterizing the model with individual state coefficients, which could lead to overfitting, particularly in cases with limited polling data for certain states. The random effect structure enhances the model’s generalizability, making it more robust when predicting support in battleground states with varying polling frequencies.

The Bayesian approach offers substantial advantages for our model by allowing us to incorporate prior information, which enhances the stability and reliability of our estimates. Given that polling data can be sparse for certain states or dates, Bayesian inference helps stabilize parameter estimates, especially when limited data points are available. We defined weakly informative priors for the model parameters: for the intercept $\beta_0$, we specified $\beta_0 \sim \text{Normal}(0, 10)$ to provide a relatively uninformative prior that allows the data to inform the estimation of baseline log-odds; for the slope $\beta_1$, we used $\beta_1 \sim \text{Normal}(0, 5)$, reflecting reasonable expectations regarding the influence of polling trends over time.

This model structure not only allows for capturing the dynamics of voter support but also provides interpretability that aligns well with the goals of our paper. The logit link function enables us to interpret results in terms of odds or probabilities of victory, which is crucial in the election context. By constructing separate models for Trump and Harris, we can directly compare their temporal trends and state-level variations without confounding effects. Thus, the Bayesian Beta regression model with a logit link function is justified as the most suitable model for our study. It accommodates flexible non-linear trends over time, accounts for regional variation across battleground states, and delivers interpretable results concerning victory probabilities. The Bayesian framework not only stabilizes estimates in the presence of sparse data but also provides meaningful uncertainty estimates, ultimately aligning with our objective of forecasting each candidate’s probability of victory in the 2024 U.S. Presidential Election.

## Model Diagnostics {#sec-model-diagnostics}
To assess the validity and reliability of our Bayesian models for predicting voter support for Kamala Harris and Donald Trump, we conducted posterior predictive checks for both models. These checks are essential for evaluating the models' ability to replicate observed polling data based on the estimated parameters. By comparing the distributions of observed support percentages with the predicted values, we can identify discrepancies and assess the overall fit of our models.

Posterior predictive checks help verify model assumptions and provide insights into how well the models capture the underlying voter dynamics. A close alignment between observed and predicted values suggests that our models effectively reflect voter support trends in the electoral context. In contrast, significant deviations may reveal areas for model improvement or highlight limitations in our approach. This diagnostic process aims to enhance the credibility of our findings and deepen our understanding of the electoral landscape as the 2024 Presidential Election approaches

>>>>

## Measurement
Voter opinions are influenced by various factors, such as media, political events, personal beliefs, and social pressures, which can change frequently, especially near an election. Pollsters simplify these complex opinions into basic polling responses like “Trump,” “Harris,” or “Undecided,” which can obscure nuanced voter sentiment. This initial conversion may not capture voters’ internal conflicts or reservations, as respondents must choose a definitive option.

These individual responses are aggregated into percentages, treating each “Trump” or “Harris” response equally, regardless of confidence level, further smoothing out complexities. The model uses these aggregated support percentages to predict election outcomes by estimating the likelihood of each candidate winning. This step assumes that poll data accurately reflects voter behavior on Election Day, which may not always hold true, as it simplifies support into binary outcomes (predicting Trump or Harris as the winner based on higher support).

The use of “state” as a predictor assumes that all responses within a state are comparable, despite internal diversity, such as urban-rural divides or cultural differences, which may not be fully captured. Similarly, the “end_date” variable tracks changes in voter sentiment over time to capture patterns influenced by campaign events or news. However, this variable simplifies continuous shifts into discrete snapshots, assuming stability between poll dates. This may overlook rapid opinion changes following significant events. Additionally, the model presumes time’s influence on voter sentiment is consistent across states, despite varying local contexts and demographics that may respond differently to events.

@@

- **'State'**: Indicates the state where the polling took place, critical for understanding regional support dynamics.
- **'Polling Date'**: The date when the poll was conducted, which helps in analyzing trends over time.
- **'Pollster'**: The organization conducting the poll, providing insight into the reliability and methodology of the polling data.
- **'Numeric Grade'**: A score assigned to each poll based on its methodology and historical accuracy, which assists in filtering out less reliable polls.
- **'Transparency Score'**: A measure indicating how transparent the pollster is regarding their methodology, further enhancing data credibility.
- **'Candidate Name'**: Identifies the candidate being polled, specifically focusing on Donald Trump and Kamala Harris.
- **Percentage of Support (pct)**: Represents the percentage of respondents supporting each candidate, which is the primary outcome variable of interest.

@@

## Model Overview
The Bayesian Beta regression models in this paper estimate the probability of victory for Donald Trump and Kamala Harris in the 2024 U.S. Presidential Election by analyzing polling data from seven battleground states. Separate models are constructed for each candidate to estimate their support, considering polling trends over time and state-level differences in voter sentiment. This approach captures how support fluctuates leading up to Election Day and predicts potential electoral outcomes. To analyze changes in voter support over time, the models use natural splines with the **'end_date_num'** (days from the earliest poll date) as a predictor, allowing for the identification of non-linear trends influenced by campaign events and news cycles. Additionally, state-level effects are incorporated as random factors, enabling the models to account for unique political landscapes across the battleground states. This regional adjustment helps highlight variations in candidate support and enhances the accuracy of predictions. Further detailed model specifications and diagnostics are provided in [Appendix -@sec-model-details].

@@

## Model Setup

@@

## Uncertainty Beyond Data and Modeling
Beyond data and modeling, external factors such as recency bias, social desirability bias, and non-response bias can influence the election outcome. Recency bias significantly shapes public perception, with many expecting Trump to win due to his 2016 victory and narrow 2020 loss, despite lagging in polls. This bias highlights a tendency to focus on recent events while overlooking broader trends. For example, in 2012, Barack Obama outperformed polls, a fact often overshadowed by recent election narratives[@discussion-article3]. This can skew how people respond to polls and how pollsters interpret results, introducing potential polling errors that our model cannot fully account for.

Social desirability bias and non-response bias add further uncertainty, especially regarding Trump’s support base. Social stigma may deter some voters from expressing conservative support openly, leading to under-reported Trump support in polls. Additionally, Trump’s supporters often have lower social trust, making them less likely to participate in polls altogether[@discussion-article3]. If certain groups systematically avoid participation, polling data may under-represent these voters, resulting in an underestimation of Trump’s support in our model.

Harris faces potential challenges such as the Bradley effect and gender-related bias. As a Black female candidate, Harris may encounter undecided voters leaning against her due to underlying biases, similar to what Hillary Clinton faced in 2016 [@discussion-article3].  If undecided voters ultimately vote against Harris, our model may overestimate her support based on polling data alone, affecting prediction accuracy.

These biases are difficult to measure or correct, reflecting deeper societal attitudes and complexities in voter behavior. Although pollsters may attempt adjustments, these are often based on assumptions that cannot fully capture the dynamics at play. Therefore, our model’s predictions should be interpreted cautiously, acknowledging that biases may lead to unexpected shifts beyond the range of data-driven forecasts.

@@

- **Pollster Ratings and Credibility**: Polls from different organizations often vary in reliability, as some pollsters have more accurate or less biased methodologies. By integrating pollster ratings, the model could assign weights to different polls based on the historical accuracy and reputation of each organization. This adjustment would allow the model to downplay the influence of potentially biased or less credible polls, particularly in competitive battleground states, where small differences in predicted support can significantly affect the outcome. This approach would lead to a more calibrated model that accounts for poll variability and thus provides a more reliable prediction.
- **Timing and Recency of Polls**: Voter sentiment is not static, and the timing of polls can significantly impact their relevance. As the election date nears, polls generally become more reflective of the final outcome. Incorporating a temporal weighting scheme that prioritizes recent polls could enhance the model’s responsiveness to last-minute shifts in voter preferences, which often occur due to campaign events, debates, or emergent political issues. This weighting approach could also help mitigate the risk of using outdated data that no longer accurately reflects current voter sentiment.
- **Economic and Political Context**: The political landscape is shaped by various external factors, such as economic indicators (e.g., unemployment rates, inflation) and prominent political events. By incorporating economic data or major political issues, such as health care or immigration policies, the model could capture contextual variables that influence voter sentiment. Such contextual factors could be especially useful in a Bayesian framework, allowing us to adjust prior distributions based on known historical influences on voter behavior.
- **Inclusive of National/Popular votes**: Our model focuses exclusively on the seven battleground states—Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin—omitting data from solidly Democratic or Republican states and the national popular vote. This exclusion limits the model’s ability to capture broader voter sentiment across the country, restricting its capacity for a holistic prediction of the 2024 Presidential Election. While battleground states are crucial due to the Electoral College system, popular vote trends from other states can indicate national shifts that may indirectly influence battleground dynamics. For example, a nationwide trend favoring or opposing a candidate could signal political momentum or reflect impactful events resonating with voters in battleground states. Without national data, our model may overlook these broader trends, potentially affecting the accuracy of its predictions in key states.

@@

# Additional Model details {#sec-model-details}
## Model Justification - Further Details
The Bayesian Beta regression model with a logit link function is well-suited for predicting the 2024 U.S. Presidential Election due to the nature of our outcome data and analysis goals. Our primary outcome variable is the proportion of support for each candidate in seven battleground states, scaled between 0 and 1. The Beta distribution is ideal for modeling such bounded data, unlike linear regression, which assumes an unbounded outcome and risks unrealistic predictions outside the 0-100% range.

To capture non-linear trends in voter support over time, we use natural splines with the variable **'end_date_num'** (days since the earliest poll date), allowing the model to adapt to shifts due to campaign events or debates. Specifying four degrees of freedom for the splines balances flexibility in modeling gradual changes with reducing noise or overfitting. Standard logistic regression lacks this adaptability and would be less effective in capturing the evolving nature of voter sentiment. Additionally, incorporating state as a random effect enables the model to estimate state-specific baseline support levels, reflecting the distinct political dynamics of each battleground state while preserving overall trend analysis. This approach prevents overfitting by avoiding an overparameterized model, particularly useful when polling data for some states is sparse. The random effect structure thus enhances model robustness and generalizability across states with varying polling frequencies.

The Bayesian approach adds value by incorporating prior information, stabilizing parameter estimates where data are limited. We applied weakly informative priors to ensure data-driven estimation: the intercept ($\beta_0 \sim \text{Normal}(0, 10)$) and slope ($\beta_1 \sim \text{Normal}(0, 5)$). This setup supports stable inference, crucial in states or periods with sparse polling data. This model structure provides clear interpretability, essential for comparing the probability of victory for Trump and Harris. The logit link function facilitates results interpretation in terms of odds or victory probabilities, aligning with our election analysis goals. Constructing separate models for each candidate allows for direct comparisons of trends and state-level variations without confounding. Thus, the Bayesian Beta regression model with a logit link function is justified as the optimal choice. It captures non-linear temporal trends, accounts for regional variations, and delivers interpretable victory probabilities, supported by Bayesian inference that enhances reliability and reflects uncertainty in forecasts.

## Model Diagnostics {#sec-model-diagnostics}
To assess the validity and reliability of our Bayesian models for predicting voter support for Kamala Harris and Donald Trump, we conducted posterior predictive checks for both models. These checks are essential for evaluating the models' ability to replicate observed polling data based on the estimated parameters. By comparing the distributions of observed support percentages with the predicted values, we can identify discrepancies and assess the overall fit of our models. Posterior predictive checks help verify model assumptions and provide insights into how well the models capture the underlying voter dynamics. A close alignment between observed and predicted values suggests that our models effectively reflect voter support trends in the electoral context. In contrast, significant deviations may reveal areas for model improvement or highlight limitations in our approach. This diagnostic process aims to enhance the credibility of our findings and deepen our understanding of the electoral landscape as the 2024 Presidential Election approaches
